---
title: "Parsing logs"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

A log file is a character vector with a format. `hs_capture()` turns it into a
data frame with one regular expression: one row per line, one column per
capture group.

```{r setup}
library(vectorscan)
```

## One format

```{r}
lines <- c(
  "2026-09-26 10:15:02.117 [INFO]  api.server:42 - listening on :8080",
  "2026-09-26 10:15:07.901 [WARN]  db.pool:118 - slow query took 1.8s",
  "--- rotated ---",
  "2026-09-26 10:16:00.004 [ERROR] auth.jwt:77 - token expired for user 12"
)

fmt <- paste0(
  "^(?<time>\\S+ \\S+) \\[(?<level>\\w+)\\]\\s+",
  "(?<location>[^ :]+):(?<line>\\d+) - (?<text>.*)$"
)

hs_capture(fmt, lines)
```

Named groups (`(?<level>...)`) name the columns. Lines that do not match the
format, like the rotation marker, give a row of `NA`, so they are easy to
find:

```{r}
log <- hs_capture(fmt, lines)
lines[is.na(log$time)]
```

## Column types

Pass a `proto`, as for `utils::strcapture()`, to get typed columns. Its
names replace the group names:

```{r}
proto <- data.frame(
  time = character(), level = character(), location = character(),
  line = integer(), text = character()
)
log <- hs_capture(fmt, lines, proto)
log$time <- as.POSIXct(log$time, format = "%Y-%m-%d %H:%M:%OS", tz = "UTC")
str(log)
```

## Reading a file

For a real file, read the lines and compile the pattern once:

```{r, eval = FALSE}
compiled <- hs_capture_compile(fmt)
log <- hs_capture(compiled, readLines("app.log"), proto)
```

## Only some lines

When only a few lines matter, filter first with `hs_detect()`, which checks
every line against many patterns in one pass on Vectorscan, then capture
fields from the lines it keeps:

```{r}
interesting <- c(error = "\\[ERROR\\]", slow = "slow query took")
keep <- hs_detect(interesting, lines)
hs_capture(fmt, lines[keep], proto)
```

## How fast

On a 1,000,000-line log (72 MB) in the format above, on an Apple M-series
laptop, each timed in a fresh R session (script in `tools/bench-capture/` in
the source repository):

| | every line matches | 1% of lines match |
|---|---|---|
| `utils::strcapture(perl = TRUE)` | 12.8 s | 6.5 s |
| `hs_capture()` | 0.60 s | 0.24 s |

That is 21 and 27 times faster. (Timing both in one session flatters
whichever runs second: R caches strings globally, and the second run finds
most of its 5 million field values already made.)

The two return identical data frames. The difference is not the regex
engine: both use PCRE2. `strcapture()` goes through `regexec()` and
`regmatches()`, which build R objects for every line. `hs_capture()` matches
and fills the columns directly in C, with PCRE2's just-in-time compiler.

## Differences from `strcapture()`

- A group that does not take part in the match (an optional group, or a
  branch not taken) is `NA` in `hs_capture()` and `""` in `strcapture()`.
- Lines PCRE2 cannot match at all, such as invalid UTF-8, give an `NA` row
  and one warning with their count.
- The pattern is always a Perl-compatible regular expression, as
  `strcapture(perl = TRUE)`.
